Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}

### 2015-01-28 inst/doc/article.Rnw
version 1.0.0

%\VignetteIndexEntry{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol}
%\VignetteKeyword{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}

@@

%% almost as usual
\author{Gregor Kastner\\WU Vienna University of Economics and Business}
\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
%\Shorttitle{The \proglang{R} package \pkg{stochvol} (DRAFT VERSION)} %% a short title (if necessary)
\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting

@@

%% an abstract and keywords
\Abstract{
The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modeling within the framework of stochastic volatility.
It utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables which can then be used for predicting future volatilities. The package can straightforwardly be employed as a stand-alone tool; moreover, it allows for easy incorporation into other MCMC samplers. The main focus of this paper is to show the functionality of \pkg{stochvol}. In addition, it provides a brief mathematical description of the model, an overview of the sampling schemes used, and several illustrative examples using exchange rate data.
}
\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
%% at least one keyword must be supplied

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Gregor Kastner\\
  Institute for Statistics and Mathematics\\
  Department of Finance, Accounting and Statistics\\
  WU Vienna University of Economics and Business\\
  Welthandelsplatz 1\\
  1020 Vienna, Austria\\
  E-mail: \email{gregor.kastner@wu.ac.at}
%  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

@@

One approach is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models. After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and applied to a vast amount of real-world problems. As an alternative, \cite{tay:fin} proposes in his seminal work to model the volatility probabilistically, i.e., through a state-space model where the logarithm of the squared volatilities -- the latent states -- follow an autoregressive process of order one. Over time, this specification became known as the \emph{stochastic volatility (SV)} model. Even though several papers \citep[e.g.,][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} provide early evidence in favor of using SV, these models have found comparably little use in applied work. This obvious discrepancy is discussed in \cite{bos:rel} who points out two reasons: the variety (and potential incompatibility) of estimation methods for SV models -- whereas the many variants of the GARCH model have basically a single estimation method -- and the lack of standard packages implementing these methods.

@@

 In \cite{kas-fru:anc}, the former issue is thoroughly investigated and an efficient MCMC estimation scheme
%using techniques such as sampling AWOL \citep[``all without a loop'', adapted from][]{rue:fas, mcc-etal:sim} and ASIS \citep[``ancillarity-sufficiency interweaving strategy'', adapted from][]{yu-men:cen}, an
is proposed. The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} are crafted to cope with the latter problem: the apparent lack of ready-to-use packages for efficiently estimating SV models.

@@

The SV model (in its centered parameterization) is thus given through
\begin{eqnarray}  
 \label{c1}
 y_{t}|\hsv_t &\sim&  \Normal{0, \exp{\hsv_{t}}},%\error_{t},% \qquad \error_{t} \sim \Normal{0,1},
 \\
 \label{c2}
 \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapartwo},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
 \label{c3}
\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapartwo/(1-\phi^2)},
\end{eqnarray}

where $\Normal{\mu, \sigmapartwo}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapartwo$. We refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} of log-variance $\mupar$, the \emph{persistence} of log-variance $\phipar$, and the \emph{volatility} of log-variance $\sigmapar$. The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{n})$ appearing in state equations (\ref{c2}) and (\ref{c3}) is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely, the log-variance process). Note that the initial state $\hsv_0$ is distributed according to the stationary distribution of the autoregressive process of order one.

\subsection{Prior distributions}
\label{priors}

@@

\subsection{MCMC Sampling}

@@

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of a variant of the ``ancillarity-sufficiency interweaving strategy'' (ASIS) which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This is commonly referred to as a reparameterization issue and dealt with extensively in scholarly literature; for an early reference see, e.g., \cite{hil-smi:par}.
In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility of log-variance $\sigmapar$ from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For more details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

@@

The usual stand-alone approach to fitting SV models with \pkg{stochvol} exhibits the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results. All these steps are described in more detail below, along with a worked example. For a stepwise incorporation of SV effects into other MCMC samplers, see Section \ref{other}.

@@

<<usd1, fig=TRUE, include=FALSE, width=11, height=6>>=
set.seed(123)
library("stochvol")
data("exrates")
ret <- logret(exrates$USD, demean = TRUE)
par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log returns")
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd1}
 \caption{Visualization of EUR-USD exchange rates included in the \pkg{stochvol} package.}
 \label{fig1}
\end{center}
\end{figure}

@@

For specifying the size of the burn-in, the parameter \code{burnin} is provided. It is the amount of MCMC iterations that are run but discarded to ensure convergence to the stationary distribution of the chain. The current default value for this parameter is $1000$ which has turned out to suffice in most situations. Nevertheless, the user is encouraged to check convergence carefully; see Section \ref{check} for more details. The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$. Consequently, the sampler is run for a total of \code{burnin + draws} iterations.

@@

<<eval=FALSE>>=
res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
                priorsigma = .1)
@

<<results=hide, echo=FALSE>>=
if (usePreCalcResultsExtra) {
 load('vignette_sampling_draws1.RData')
} else {
 res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
 #save(res, file = 'vignette_sampling_draws1.RData')
}
@

\begin{CodeOutput}
Calling GIS_C MCMC sampler with 11000 iter. Series length is 3139.

@@

Timing (elapsed): 17.935 seconds.
613 iterations per second.

Converting results to coda objects... Done!
Summarizing posterior draws... Done!
\end{CodeOutput}

@@

There are several plotting functions specifically designed for objects of class \code{svsample} which are described in the following paragraphs.
 \begin{enumerate}
  \item[(1)] \code{volplot}: Plots posterior quantiles of the latent volatilities in percent, i.e., $100\exp(h_t/2)$, over time. Apart from the mandatory \code{svsample}-object itself, this function takes several optional arguments. Only some are mentioned here; for an exhaustive list please see the corresponding help document accessible through \code{?volplot} or \code{help(volplot)}. Selected optional arguments that are commonly used include \code{forecast} for $n$-step-ahead volatility prediction, \code{dates} for labels on the $x$-axis, alongside some graphical parameters. The code snipped below shows a typical example and Figure~\ref{fig3} displays its output.

@@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd3}
 \caption{Visualization of estimated contemporaneous volatilities of EUR-USD exchange rates, as provided by \code{volplot}. If not specified otherwise, posterior medians and 5\%/95\% quantiles are plotted. The dotted lines on the right indicate predicted future volatilities.}
 \label{fig3}
\end{center}
\end{figure}

@@

<<usd4, fig=TRUE, include=FALSE, width=11, height=7>>=
res <- updatesummary(res, quantiles = c(.01, .1, .5, .9, .99))
volplot(res, forecast = 100, dates = exrates$date[-1])
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.96\textwidth]{article-usd4}
 \caption{As above, now with medians (black line) and 1\%/10\%/90\%/99\% quantiles (gray lines). This behavior can be achieved through a preceding call of \code{updatesummary}.}
 \label{fig4}
\end{center}
\end{figure}

@@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.96\textwidth]{article-usd5}
 \caption{Trace plots of posterior draws for the parameters $\mupar, \phipar, \sigmapar$.}
 %The underlying dataset is again EUR-USD exchange rates.}
 \label{fig5}
\end{center}
 \end{figure}

@@

In \proglang{R}, this can straightforwardly be coded as follows:
\begin{itemize}
 \item Set the seed to make results reproducible and simulate some data from the underlying model:
<<>>=
set.seed(123456)
n <- 1000
beta.true <- c(.1, .5)
sigma.true <- 0.01
X <- matrix(c(rep(1, n), rnorm(n, sd = sigma.true)), nrow = n)
y <- rnorm(n, X %*% beta.true, sigma.true)
@

\item Specify the size of the burn-in and the number of draws thereafter; set the prior parameters:
<<>>=
burnin <- 100
draws <- 5000
b0 <- matrix(c(0, 0), nrow = ncol(X))
B0inv <- diag(c(10^-10, 10^-10))
c0 <- 0.001
C0 <- 0.001
@
 \item Pre-calculate some values outside the main MCMC loop:
<<>>=
p <- ncol(X)
preCov <- solve(crossprod(X) + B0inv)
preMean <- preCov %*% (crossprod(X, y) + B0inv %*% b0)
preDf <- c0 + n/2 + p/2
@
\item Assign some storage space for holding the draws and set an initial value for $\sigmaepsextrasuper2$:
<<>>=
draws1 <- matrix(NA_real_, nrow = draws, ncol = p + 1)
colnames(draws1) <- c(paste("beta", 0:(p-1), sep='_'), "sigma")
sigma2draw <- 1
@
\item Run the main sampler: Iteratively draw from the conditional bivariate Gaussian distribution $\betab|\yb,\sigmaepsextrasuper2$, e.g., through the use of \pkg{mvtnorm} \citep{r:mvt}, and the conditional Inverse Gamma distribution $\sigmaepsextrasuper2|\yb,\betab$.
<<eval=FALSE>>=
for (i in -(burnin-1):draws) {
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean,
					 sigma2draw * preCov))
 tmp <- C0 + .5 * (crossprod(y - X %*% betadraw) +
 		  crossprod((betadraw - b0), B0inv) %*%
		   (betadraw - b0))
 sigma2draw <- 1 / rgamma(1, preDf, rate = tmp)
 if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
}
@
<<echo=FALSE>>=
for (i in -(burnin-1):draws) {
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean, sigma2draw*preCov))
 tmp <- C0 + .5*(crossprod(y - X%*%betadraw) +
 		 crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
 sigma2draw <- 1/rgamma(1, preDf, rate = tmp)
 if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
}
@
\item Finally, summarize and visualize the posterior draws:
<<>>=
colMeans(draws1)
@

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=.98\textwidth]{article-homo}
\caption{Trace plots and kernel density estimates for some draws from the marginal posterior distributions in the regression model with heteroskedastic errors. Underlying data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $\sigmaepsextrasuper{\text{true}} = 0.01$, $n=1000$.}
\label{homodraws}
\end{center}
\end{figure}

@@

Here is how it goes:
\begin{itemize}
 \item Simulate some data:
<<>>=
mu.true <- log(sigma.true^2)
phi.true <- 0.97
vv.true <- 0.3
simresid <- svsim(n, mu = mu.true, phi = phi.true, sigma = vv.true)
y <- X %*% beta.true + simresid$y
@
 \item Specify configuration parameters and prior values:
<<>>=
draws <- 50000
burnin <- 1000
thinning <- 10
priormu <- c(-10, 2)
priorphi <- c(20, 1.5)
priorsigma <- 1
@
 \item Assign some storage space for holding the draws and set initial values:
<<>>=
draws2 <- matrix(NA_real_, nrow = floor(draws / thinning),
		 ncol = 3 + n + p)
colnames(draws2) <- c("mu", "phi", "sigma",
		      paste("beta", 0:(p-1), sep = '_'),
                      paste("h", 1:n, sep='_'))
betadraw <- c(0, 0)
svdraw <- list(para = c(mu = -10, phi = .9, sigma = .2),
	       latent = rep(-10, n))
@
 \item Run the main sampler, i.e., iteratively draw
  \begin{itemize}
   \item the latent volatilities/parameters by conditioning on the regression parameters and calling \code{svsample2},
   \item the regression parameters by conditioning on the latent volatilities and calling \code{rmvnorm}:
  \end{itemize}
<<eval=FALSE>>=
for (i in -(burnin-1):draws) {
 ytilde <- y - X %*% betadraw
 svdraw <- svsample2(ytilde, startpara = para(svdraw),
                     startlatent = latent(svdraw), priormu = priormu,
                     priorphi = priorphi, priorsigma = priorsigma)
 normalizer <- as.numeric(exp(-latent(svdraw)/2))
 Xnew <- X * normalizer
 ynew <- y * normalizer
 Sigma <- solve(crossprod(Xnew) + B0inv)
 mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))
 if (i > 0 & i %% thinning == 0) { 
  draws2[i/thinning, 1:3] <- para(svdraw)
  draws2[i/thinning, 4:5] <- betadraw
  draws2[i/thinning, 6:(n+5)] <- latent(svdraw)
 }
}
@

<<echo=FALSE>>=
if (usePreCalcResults) {
 load('vignette_sampling_draws2.RData')
} else {
for (i in -(burnin-1):draws) {

@@

 # store the results:
 if (i > 0 & i %% thinning == 0) { 
  draws2[i/thinning, 1:3] <- para(svdraw)
  draws2[i/thinning, 4:5] <- betadraw
  draws2[i/thinning, 6:(n+5)] <- latent(svdraw)
 }
}
 draws2selection <- draws2[,4:8]
 #save(draws2selection, file = 'vignette_sampling_draws2.RData')
}
@
\item Finally, summarize and visualize (some) posterior draws:
<<eval=FALSE>>=
colMeans(draws2[,4:8])
@

@@

<<eval=FALSE>>=
plot(coda::mcmc(draws2[,4:7]))
@

>>>>

%\VignetteIndexEntry{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol}
%\VignetteKeyword{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}

@@

%% almost as usual
\author{Gregor Kastner\\WU Vienna University of Economics and Business}
\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
%\Shorttitle{The \proglang{R} package \pkg{stochvol} (DRAFT VERSION)} %% a short title (if necessary)
\Plaintitle{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting

@@

%% an abstract and keywords
\Abstract{
The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modeling within the framework of stochastic volatility.
It utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables which can then be used for predicting future volatilities. The package can straightforwardly be employed as a stand-alone tool; moreover, it allows for easy incorporation into other MCMC samplers. The main focus of this paper is to show the functionality of \pkg{stochvol}. In addition, it provides a brief mathematical description of the model, an overview of the sampling schemes used, and several illustrative examples using exchange rate data.
}
\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
%\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
%% at least one keyword must be supplied

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Gregor Kastner\\
  Institute for Statistics and Mathematics\\
  Department of Finance, Accounting and Statistics\\
  WU Vienna University of Economics and Business\\
%  Building D4, Level 4\\
  Welthandelsplatz 1\\
  1020 Vienna, Austria\\
%  Telephone: +43/1/31336-5593\\
%  Fax: +43/1/31336-90-5593\\
  E-mail: \email{gregor.kastner@wu.ac.at}\\
  URL: \url{http://www.wu.ac.at/statmath/en/faculty_staff/faculty/gkastner/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

\section*{Preface}
This vignette corresponds to a forthcoming article of the same name in the Journal of Statistical Software. At the time of this writing, the vignette and the published article coincide; nevertheless, this version might receive minor updates over time. To cite, please use \cite{kas:dea}. Further information about citing \pkg{stochvol} can be obtained in \proglang{R} by installing the package, e.g., through \code{install.packages("stochvol")}, and typing \code{citation("stochvol")}.

@@

One approach is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models. After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and applied to a vast amount of real-world problems. As an alternative, \cite{tay:fin} proposes in his seminal work to model the volatility probabilistically, i.e., through a state-space model where the logarithm of the squared volatilities -- the latent states -- follow an autoregressive process of order one. Over time, this specification became known as the \emph{stochastic volatility (SV)} model. Even though several papers \citep[e.g.,][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} provide early evidence in favor of using SV, these models have found comparably little use in applied work. This obvious discrepancy is discussed in \cite{bos:rel} who points out two reasons: the variety (and potential incompatibility) of estimation methods for SV models -- whereas the many variants of the GARCH model have basically a single estimation method -- and the lack of standard software packages implementing these methods.

@@

 In \cite{kas-fru:anc}, the former issue is thoroughly investigated and an efficient MCMC estimation scheme
%using techniques such as sampling AWOL \citep[``all without a loop'', adapted from][]{rue:fas, mcc-etal:sim} and ASIS \citep[``ancillarity-sufficiency interweaving strategy'', adapted from][]{yu-men:cen}, an
is proposed. The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} are crafted to cope with the latter problem: the apparent lack of ready-to-use software packages for efficiently estimating SV models.

@@

The SV model can thus be conveniently expressed in hierarchical form. In its centered parameterization, it is given through
\begin{eqnarray}  
 \label{c1}
 y_{t}|\hsv_t &\sim&  \Normal{0, \exp{\hsv_{t}}},%\error_{t},% \qquad \error_{t} \sim \Normal{0,1},
 \\
 \label{c2}
 \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapartwo},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
 \label{c3}
\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapartwo/(1-\phi^2)},
\end{eqnarray}

where $\Normal{\mu, \sigmapartwo}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapartwo$. We refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} of log-variance $\mupar$, the \emph{persistence} of log-variance $\phipar$, and the \emph{volatility} of log-variance $\sigmapar$. The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{n})$ appearing in Equation~\ref{c2} and Equation~\ref{c3} is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely, the log-variance process). Note that the initial state $\hsv_0$ appearing in Equation~\ref{c3} is distributed according to the stationary distribution of the autoregressive process of order one.

\subsection{Prior distribution}
\label{priors}

@@

\subsection{MCMC sampling}

@@

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of a variant of the ``ancillarity-sufficiency interweaving strategy'' (ASIS) which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This is commonly referred to as a reparameterization issue and dealt with extensively in scholarly literature; for an early reference see, e.g., \cite{hil-smi:par}.
For the model at hand, a move of this kind can be achieved by transferring the level of log-variance $\mupar$ and/or the volatility of log-variance $\sigmapar$ from the state process (Equations~\ref{c2}~and~\ref{c3}) to the observation process (Equation~\ref{c1}) through a simple reparameterization of $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For more details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

@@

The usual stand-alone approach to fitting SV models with \pkg{stochvol} exhibits the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results. All these steps are described in more detail below, along with a worked example. For a stepwise incorporation of SV effects into other MCMC samplers, see Section~\ref{other}.

@@

<<usd1, fig=TRUE, include=FALSE, width=11, height=6>>=
set.seed(123)
library("stochvol")
data("exrates")
ret <- logret(exrates$USD, demean = TRUE)
par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, 0.5), mgp = c(2, 0.6, 0))
plot(exrates$date, exrates$USD, type = "l",
  main = "Price of 1 EUR in USD")
plot(exrates$date[-1], ret, type = "l", main = "Demeaned log returns")
@
\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd1}
 \caption{Visualization of EUR-USD exchange rates included in the \pkg{stochvol} package.}
 \label{fig1}
\end{center}
\end{figure}

@@

For specifying the size of the burn-in, the parameter \code{burnin} is provided. It is the amount of MCMC iterations that are run but discarded to ensure convergence to the stationary distribution of the chain. The current default value for this parameter is $1000$ which has turned out to suffice in most situations. Nevertheless, the user is encouraged to check convergence carefully; see Section~\ref{check} for more details. The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$. Consequently, the sampler is run for a total of \code{burnin + draws} iterations.

@@

<<eval=FALSE>>=
res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
  priorsigma = 0.1)
@

<<results=hide, echo=FALSE>>=
if (usePreCalcResultsExtra) {
 load("vignette_sampling_draws1.RData")
} else {
 res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
 #save(res, file = "vignette_sampling_draws1.RData")
}
@

\begin{Soutput}
Calling GIS_C MCMC sampler with 11000 iter. Series length is 3139.

@@

Timing (elapsed): 12.92 seconds.
851 iterations per second.

Converting results to coda objects... Done!
Summarizing posterior draws... Done!
\end{Soutput}

@@

There are several plotting functions specifically designed for objects of class \code{svsample} which are described in the following paragraphs.
 \begin{enumerate}
  \item[(1)] \code{volplot}: Plots posterior quantiles of the latent volatilities in percent, i.e., empirical quantiles of the posterior distribution of $100\exp(h_t/2)$, over time. Apart from the mandatory \code{svsample}-object itself, this function takes several optional arguments. Only some are mentioned here; for an exhaustive list please see the corresponding help document accessible through \code{?volplot} or \code{help(volplot)}. Selected optional arguments that are commonly used include \code{forecast} for $n$-step-ahead volatility prediction, \code{dates} for labels on the $x$-axis, alongside some graphical parameters. The code snipped below shows a typical example and Figure~\ref{fig3} displays its output.

@@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd3}
 \caption{Visualization of estimated contemporaneous volatilities of EUR-USD exchange rates, as provided by \code{volplot}. If not specified otherwise, posterior medians and 5\%/95\% quantiles are plotted. The dotted lines on the right indicate predicted future volatilities.}
 \label{fig3}
\end{center}
\end{figure}

@@

<<usd4, fig=TRUE, include=FALSE, width=11, height=7>>=
res <- updatesummary(res, quantiles = c(0.01, 0.1, 0.5, 0.9, 0.99))
volplot(res, forecast = 100, dates = exrates$date[-1])
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd4}
 \caption{As above, now with medians (black line) and 1\%/10\%/90\%/99\% quantiles (gray lines). This behavior can be achieved through a preceding call of \code{updatesummary}.}
 \label{fig4}
\end{center}
\end{figure}

@@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd5}
 \caption{Trace plots of posterior draws for the parameters $\mupar, \phipar, \sigmapar$.}
 %The underlying dataset is again EUR-USD exchange rates.}
 \label{fig5}
\end{center}
 \end{figure}

@@

In \proglang{R}, this can straightforwardly be coded as follows:
\begin{itemize}
 \item Set the seed to make results reproducible and simulate some data from the underlying model:
<<>>=
set.seed(123456)
n <- 1000
beta.true <- c(0.1, 0.5)
sigma.true <- 0.01
X <- matrix(c(rep(1, n), rnorm(n, sd = sigma.true)), nrow = n)
y <- rnorm(n, X %*% beta.true, sigma.true)
@

\item Specify the size of the burn-in and the number of draws thereafter; set the prior parameters:
<<>>=
burnin <- 100
draws <- 5000
b0 <- matrix(c(0, 0), nrow = ncol(X))
B0inv <- diag(c(10^-10, 10^-10))
c0 <- 0.001
C0 <- 0.001
@
 \item Pre-calculate some values outside the main MCMC loop:
<<>>=
p <- ncol(X)
preCov <- solve(crossprod(X) + B0inv)
preMean <- preCov %*% (crossprod(X, y) + B0inv %*% b0)
preDf <- c0 + n/2 + p/2
@
\item Assign some storage space for holding the draws and set an initial value for $\sigmaepsextrasuper2$:
<<>>=
draws1 <- matrix(NA_real_, nrow = draws, ncol = p + 1)
colnames(draws1) <- c(paste("beta", 0:(p-1), sep = "_"), "sigma")
sigma2draw <- 1
@
\item Run the main sampler: Iteratively draw from the conditional bivariate Gaussian distribution $\betab|\yb,\sigmaepsextrasuper2$, e.g., through the use of \pkg{mvtnorm} \citep{r:mvt}, and the conditional Inverse Gamma distribution $\sigmaepsextrasuper2|\yb,\betab$.
<<>>=
for (i in -(burnin-1):draws) {
  betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean,
    sigma2draw * preCov))
  tmp <- C0 + 0.5 * (crossprod(y - X %*% betadraw) +
    crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
  sigma2draw <- 1 / rgamma(1, preDf, rate = tmp)
  if (i > 0) draws1[i, ] <- c(betadraw, sqrt(sigma2draw))
}
@
\item Calculate posterior means in order to obtain point estimates for the parameters:
<<>>=
colMeans(draws1)
@

\item Visualize the draws through \pkg{coda}'s native \code{plot} method:

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{article-homo}
\caption{Trace plots and kernel density estimates for some draws from the marginal posterior distributions in the regression model with heteroskedastic errors. Underlying data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $\sigmaepsextrasuper{\text{true}} = 0.01$, $n=1000$.}
\label{homodraws}
\end{center}
\end{figure}

@@

Here is how it goes:
\begin{itemize}
 \item Simulate some data:
<<>>=
mu.true <- log(sigma.true^2)
phi.true <- 0.97
vv.true <- 0.3
simresid <- svsim(n, mu = mu.true, phi = phi.true, sigma = vv.true)
y <- X %*% beta.true + simresid$y
@
 \item Specify configuration parameters and prior values:
<<>>=
draws <- 50000
burnin <- 1000
thinning <- 10
priormu <- c(-10, 2)
priorphi <- c(20, 1.5)
priorsigma <- 1
@
 \item Assign some storage space for holding the draws and set initial values:
<<>>=
draws2 <- matrix(NA_real_, nrow = floor(draws / thinning),
  ncol = 3 + n + p)
colnames(draws2) <- c("mu", "phi", "sigma",
  paste("beta", 0:(p-1), sep = "_"), paste("h", 1:n, sep = "_"))
betadraw <- c(0, 0)
svdraw <- list(para = c(mu = -10, phi = 0.9, sigma = 0.2),
  latent = rep(-10, n))
@
 \item Run the main sampler, i.e., iteratively draw
  \begin{itemize}
   \item the latent volatilities/parameters by conditioning on the regression parameters and calling \code{svsample2},
   \item the regression parameters by conditioning on the latent volatilities and calling \code{rmvnorm}:
  \end{itemize}
<<eval=FALSE>>=
for (i in -(burnin-1):draws) {
  ytilde <- y - X %*% betadraw
  svdraw <- svsample2(ytilde, startpara = para(svdraw),
    startlatent = latent(svdraw), priormu = priormu,
    priorphi = priorphi, priorsigma = priorsigma)
  normalizer <- as.numeric(exp(-latent(svdraw) / 2))
  Xnew <- X * normalizer
  ynew <- y * normalizer
  Sigma <- solve(crossprod(Xnew) + B0inv)
  mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
  betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))
  if (i > 0 & i %% thinning == 0) { 
    draws2[i/thinning, 1:3] <- para(svdraw)
    draws2[i/thinning, 4:5] <- betadraw
    draws2[i/thinning, 6:(n+5)] <- latent(svdraw)
  }
}
@

<<echo=FALSE>>=
if (usePreCalcResults) {
 load("vignette_sampling_draws2.RData")
} else {
for (i in -(burnin-1):draws) {

@@

 # store the results:
 if (i > 0 & i %% thinning == 0) { 
  draws2[i/thinning, 1:3] <- para(svdraw)
  draws2[i/thinning, 4:5] <- betadraw
  draws2[i/thinning, 6:(n+5)] <- latent(svdraw)
 }
}
 draws2selection <- draws2[,4:8]
 #save(draws2selection, file = 'vignette_sampling_draws2.RData')
}
@
\item Finally, visualize and summarize (some) posterior draws:

<<eval=FALSE>>=
plot(coda::mcmc(draws2[, 4:7]))
@

<<hetero, fig=TRUE, include=FALSE, echo=FALSE, results=hide, width=7.5, height=8>>=
par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
plot(coda::mcmc(draws2selection[, 1:4]))
@

\begin{figure}[!ht]
 \begin{center}
\includegraphics[width=\textwidth]{article-hetero}
\caption{Trace plots and kernel density estimates
%for draws from the marginal posterior distributions of the parameters
in the regression model with heteroskedastic errors. Data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $h_1^\text{true} = -8.28$, $h_2^\text{true} = -8.50$, $n=1000$.}
\label{heterodraws}
 \end{center}
\end{figure}

<<eval=FALSE>>=
colMeans(draws2[, 4:8])
@

@@
